---
title: "Coursera, Data Science, Reproducible Research, Assessment1"
output: html_document
---
### Installing or loading packages.

Chron is for easily determining whether a date is on a weekend or not.
```{r, echo=TRUE}
if(!require(chron)){
    install.packages("chron")
    library(chron)
}

if(!require(lattice)){
    install.packages("lattice")
    library(lattice)
}
```
## Loading and preprocessing the data

Unzipping the file.
```{r, echo=TRUE}
unzip("./activity.zip", overwrite = TRUE)
```

Assigning the *.csv-file* to **activity_data**.
```{r, echo=TRUE}
activity_data <- read.csv("./activity.csv")
```

Formatting the date as date class.
```{r, echo=TRUE}
activity_data$date <- as.Date(activity_data$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

Aggregating the total steps by days.
```{r, echo=TRUE}
daily_steps <- aggregate(steps ~ date, data = activity_data, sum, na.rm = TRUE)
```

Creating the histogram.
```{r, echo=TRUE}
hist(daily_steps$steps, xlab = "Daily Steps", main = "Histogram of number of steps per day", col = "green")
```

The mean number of steps taken per day is:
```{r, echo=TRUE}
mean(daily_steps$steps)
```

The median number of steps taken per day is:
```{r, echo=TRUE}
median(daily_steps$steps)
```

## What is the average daily activity pattern?

Calculating the average number of steps in an interval.
```{r, echo=TRUE}
interval_mean <- aggregate(steps ~ interval, data = activity_data, mean, na.rm = TRUE)
```

Generating the time series plot.
```{r, echo=TRUE}
plot(x = interval_mean$interval, y = interval_mean$steps, type = "l", main = "Average number of steps per interval", xlab = "Interval", ylab = "Average number of steps", col = "blue")
```

The 5-minute interval which contains the maximum average number of steps.
```{r, echo=TRUE}
interval_mean[which.max(interval_mean$steps),]
```

## Imputing missing values.

The number of missing values in the data set.
```{r, echo=TRUE}
missings <- sum(is.na(activity_data$steps))
```

The number of missings is `r missings`.

Replacing missing values with the average value from the same interval. Imputed data set created.

```{r, echo=TRUE}
imputed <- numeric()

for (i in 1:nrow(activity_data)) {
    row <- activity_data[i, ]
    if (is.na(row$steps)) {
      filler <- subset(interval_mean, interval == row$interval)$steps
    } else {
      filler <- row$steps
    }
    imputed <- c(imputed, filler)
}

activity_imputed <- activity_data
activity_imputed$steps <- imputed
```

Imputed data set is used to summarize the total number of steps by day.
```{r, echo=TRUE}
daily_steps_imputed <- aggregate(steps ~ date, data = activity_imputed, sum, na.rm = TRUE)
```

Imputed data set is used to repeat the process to create the histogram.
```{r, echo=TRUE}
hist(daily_steps_imputed$steps, xlab = "Daily Steps", main = "Histogram of number of steps per day (imputed data)", col = "blue")
```

The mean number of steps taken per day with the imputed data is:
```{r, echo=TRUE}
mean(daily_steps_imputed$steps)
```

The median number of steps taken per day with the imputed data is:
```{r, echo=TRUE}
median(daily_steps_imputed$steps)
```


## Are there differences in activity patterns between weekdays and weekends?

The mean and median are now identical and both have the value of the unimputed mean.
```{r, echo=TRUE}
activity_imputed$weekend <- is.weekend(activity_imputed$date)
class(activity_imputed$weekend)
```

Averaging the number of steps by interval and whether its a weekend or not.
```{r, echo=TRUE}
interval_mean_imputed <- aggregate(steps ~ interval + weekend, data = activity_imputed, mean, na.rm = TRUE)
head(interval_mean_imputed)
```

Creating the plot, weekend on top, weekdays at the bottom.
```{r, echo=TRUE}
xyplot(steps ~ interval | weekend, interval_mean_imputed, type = "l", layout = c(1, 2), 
    xlab = "Interval", ylab = "Number of steps per interval", main = "Number of steps per interval, weekend (TRUE) versus weekdays (FALSE)")
```

It seems that the step activity spikes in the morning during weekdays and is more evenly distributed during weekends.